Thesis Title

### 2022-11-04 chapters/chapter_6.tex
finish hAMRonization paper, fix names formatting on papers

C. I. Mendes, F. Maguire, A. Manuele, D. Fornika, S. Tausch, T. Le-Viet, J. Phelan, C. Meehan, A. Raphenya, B. Alcock, E. Culp, A. McArthur, M. Feldgarden, G. Tyson, M. Galas, J. Campos, A. Witney, D. Aanensen, A. Black, L. Katz, P. Oluniyi, I. Olawoye, R. Timme, A. T. Vasconcelos, A. Page, D. MacCannell, E. Griffiths. 
hAMRonization: Enhancing antimicrobial resistance prediction using PHA4GE standards and specification.

@@

Catarina I Mendes$^{1,*}$, 
E J Griffiths$^{2,*}$, 
A Manuele, 
D Fornika, 
S H Tausch, 
T Le-Viet, 
A R Raphenya, 
B Alcock, 
E Culp, 
A G McArthur, 
M Feldgarden, 
G H Tyson, 
M Galas, 
J Campos, 
A A Witney, 
D M Aanensen, 
A Black, 
E Hodcroft, 
L S Katz, 
P E Oluniyi, 
I B Olawoye, 
R E Timme, 
A T R Vasconcelos, 
A J Page, 
D R MacCannell, 
F Maguire
on behalf of the Public Health Alliance for Genomic Epidemiology (PHA4GE) consortium Data Structures Working Group

@@

@@

The PHA4GE hAMRonization Detection specification was designed to harmonize information prioritised by public health practitioners for genomic surveillance, outbreak investigations and research with direct public health impacts. Fields from various widely used AMR detection tools were compared, and a set of 36 fields were selected to capture information regarding the software and database provenance, information about reference genes and genomes, as well as mutations and resistance genes detected in sequences of interest. The field labels are standardised using terms sourced or contributed to open-source ontologies. Ontologies are sets of controlled vocabulary arranged in a hierarchy, in which terms are linked using logical relationships and the meanings of the terms are disambiguated by the assignment of unique and persistent identifiers. The positions of detected mutations with respect to reference genes/genomes/proteins are specified using the machine-readable Human Genome Variation Society (HGVS) notation system, a sequence variant nomenclature that has been used previously for encoding tuberculosis drug resistance mutations (den Dunnen et al., 2016). To translate HGVS notations into clinician-friendly, human-readable language, a simplified interpretation of mutation coordinates is also provided within the specification. A list of the fields in the specification, along with their ontology identifiers, definitions, expected value types, examples of use, and additional guidance for implementation, are presented in Table 1. It should be noted that the guidance is meant to provide recommendations for developers and may contain suggestions that go beyond the immediate aims of the specification.

@@

To enable automated data transformations from variable tool outputs to the PHA4GE hAMRonization specification, fields from 18 gene and mutation detection tools (i.e. ABRicate, AMRFinderPlus, AMRPlusPlus, ARIBA, C-SSTAR, DeepARG, GROOT, KmerResistance, ResFams, ResFinder (includes PointFinder), SraX, SRST2, and StarAMR) were mapped to the specification (see example mappings in Supplementary Figure 1). It’s worth noting that not all tools provide values for all specification fields. Also, many tools provide outputs beyond what is captured by the specification.

Parsers were then constructed to automate the transformation and movement of outputs into the appropriate fields in the hAMRonization reports. Parser code is open source, and all code and installation instructions are available at \url{https://github.com/pha4ge/hAMRonization}. hAMRonization can be executed from the command-line, and alternatively, can be installed and used in Galaxy via the Galaxy tool shed.
hAMRonization provides both tabular and interactive reporting formats. The interactive report provides the user with an HTML file that can be opened within the browser for navigable data exploration.

@@

The developed  hAMRonization snakemake pipeline (see Materials and  Methods;  hAMRonization Proof of Concept Workflow) was used to run two sets of data through a selection of AMR detection tools, using various AMR gene databases.

@@

The ReLAVRA pilot included National Reference Laboratories from Argentina, Brasil, Chile, Colombia, Costa Rica, México and Paraguay. Laboratories and were provided with a dataset of 20 \textit{Shigella sonnei} genomes from a previous Latin America regional project analysed with 5 different AMR detection tools: ABRIcate, AMRFinderPlus, ARIBA, SRST2, StarAMR. Participants received the outputs of the AMR detection tools together with the instructions for installing and running hAMRonization. When participants completed the exercise, they were asked to complete the structured poll. Twelve responses from the seven participating nations were received, and a summary report was prepared by ReLAVRA and provided to PHA4GE. Most of the participants were bioinformaticians and laboratory technicians. For bioinformaticians, the installation instructions were easy to follow, however, laboratory technicians without a bioinformatics background expressed difficulties largely with the dependencies installation. Participants agreed that all harmonised output fields provided relevant information for AMR detection but the most useful were the fields that cover the AMR gene detected and the confidence of that result. Respondents were concerned about the importance of having knowledge of AMR and genomic data analysis for the correct interpretation of the results, especially with regard to decision-makers in clinical settings. An example related to this concern was that when detecting \textit{blaTEM-1} gene by AMRFinderPlus, the field “antimicrobial agent” indicates "Beta-lactam", but this type of drug includes a wide variety of antibiotics, from amoxicillin to extended-spectrum cephalosporins (ESC) and carbapenems. However, \textit{TEM-1} is an extended-spectrum beta-lactamase that does not confer resistance to ESC or carbapenems. While this concern arises from the output of AMRFinderPlus rather than hAMRonization, this challenge is a barrier to the implementation of genomics for AMR surveillance in clinical settings and should be addressed. PHA4GE is cataloguing such critical challenges and will work with the community to solve them. In terms of output summary format, most of the participants preferred the tabular report, citing that a single file combining all the relevant information was more convenient and amenable to processing with different downstream software. Some participants preferred citing its user-friendly visual interface was helpful for comparing which tools identified the same gene. Recommendations from participants can be found in Figure 2.

@@

A dataset of 89 Klebsiella pneumoniae were analysed using RGI, relying on BLASTp and the CARD database, RGI btw, relying on Bowtie2 and the CARD database, AMRFinderPlus which uses BLASTx and the BARRG database, and ResFinder with BLASTn and the ResFinder database. The aim was to determine the number of unique AMR genes identified, with a large variation having been identified across the four tools (Figure 3). RGI bwt and RGI identified the highest number of hits (n=223 and n=140 respectively), albeit using different identification methods both use the same database. AMRFinderPlus and Resfinder identified an order of magnitude less AMR genes (n=70 and n=46 respectively), even though both rely on BLAST, as does RGI. This highlights the impact of the database chosen when assessing AMR genes.

@@

A set of 87 \textit{Pseudomonas aeruginosa} sequences (ST 111) associated with a Metallo-Carbapenemases resistance outbreak in the United Kingdom \cite{witney_genome_2014}, were analysed using ABRIcate, AMRFinderplus, C-SSTAR, Resfinder, StarAMR with default parameters and databases. The resulting outputs were combined using hAMRonization (Figure 4). Most isolates (n=73) carried the \textit{VIM-2} gene, conferring the observed resistance phenotype. Some genes, such as \textit{crpP} and \textit{sul1},  are repeatedly identified by most of the tools. Differences in gene nomenclature in different databases complicate comparisons e.g. the same gene was also seen with different spelling, such as \textit{blaVIM-2} being called a \textit{blaVIM}. Nomenclature issues aside, harmonised outputs from different tools served to increase the body of evidence (and confidence in results) indicating the role of different AMR determinants in the outbreak.

@@

\subsection{hAMRonization Proof of Concept Workflow}

@@

>>>>

C. I. Mendes, E. Griffiths, A. Manuele, D. Fornika, S. Tausch, T. Le-Viet, J. Phelan, C. Meehan, A. Raphenya, B. Alcock, E. Culp, A. McArthur, M. Feldgarden, G. Tyson, M. Galas, J. Campos, A. Witney, D. Aanensen, A. Black, L. Katz, P. Oluniyi, I. Olawoye, R. Timme, A. T. Vasconcelos, A. Page, D. MacCannell, F. Maguire. 
hAMRonization: Enhancing antimicrobial resistance prediction using PHA4GE standards and specification.

@@

Catarina I. Mendes$^{1,*}$, 
Emma J. Griffiths$^{2,*}$, 
Alex Manuele$^3$, 
Dan Fornika$^4$, 
Simon H. Tausch$^5$, 
Thanh Le-Viet $^6$, 
Amogelang R. Raphenya$^7$, 
Brian Alcock$^7$, 
Elizabeth Culp$^7$, 
Andrew G. McArthur$^7$, 
Michael Feldgarden$^8$, 
Gregory H. Tyson$^9$, 
Marcelo Galas$^{10}$, 
Josefina Campos$^{11}$, 
Adam A. Witney$^{12}$, 
David M. Aanensen$^{13,14}$, 
Allison Black$^{15}$, 
Emma Hodcroft$^{16,17}$, 
Lee S. Katz$^{18,19}$, 
Paul E. Oluniyi$^{20,21}$, 
Idowu B. Olawoye$^{20,21}$, 
Ruth E. Timme$^{22}$, 
Ana Tereza R. Vasconcelos$^{23}$, 
Andrew J. Page$^6$, 
Duncan R. MacCannell$^{19}$, 
Finlay Maguire$^3$
on behalf of the Public Health Alliance for Genomic Epidemiology (PHA4GE) consortium Data Structures Working Group

@@

$^2$  Faculty of Health Sciences, Simon Fraser University, Burnaby, BC, Canada

$^3$ Faculty of Computer Science, Dalhousie University, NS, Canada

$^4$ British Columbia Centre for Disease Control, Vancouver, BC, Canada

$^5$ Department of Biological Safety, German Federal Institute for Risk Assessment, 10589 Berlin, Germany

$^6$ Quadram Institute, Norwich, UK

$^7$ Department of Biochemistry and Biomedical Sciences and the Michael G. DeGroote Institute for Infectious Disease Research, McMaster University; Hamilton; Ontario; Canada

$^8$ National Center for Biotechnology Information, National Library of Medicine, National Institutes of Health; Bethesda; Maryland; USA

$^9$ Center for Veterinary Medicine, U.S. Food and Drug Administration; Laurel; Maryland; USA

$^{10}$ Communicable Diseases and Environmental Determinants of Health, Pan American Health Organization; Washington DC; USA

$^{11}$ INEI-ANLIS "Dr Carlos G. Malbrán"; Buenos Aires;  Argentina

$^{12}$ Institute for Infection and Immunity, St George’s, University of London; London; UK

$^{13}$ Centre for Genomic Pathogen Surveillance, Wellcome Genome Campus; Cambridge; UK;

$^{14}$ The Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, Nuffield Department of Medicine, University of Oxford; Oxford; UK

$^{15}$ Department of Epidemiology, University of Washington; Washington; USA.

$^{16}$ Biozentrum, University of Basel, Basel, Switzerland

$^{17}$ Swiss Institute of Bioinformatics; Lausanne; Switzerland

$^{18}$ Center for Food Safety, University of Georgia; Georgia; USA;

$^{19}$ National Center for Emerging and Zoonotic Infectious Diseases, Centers for Disease Control and Prevention; Georgia; USA

$^{20}$ African Center of Excellence for Genomics of Infectious Diseases (ACEGID), Redeemer's University, Ede; Osun State; Nigeria;

$^{21}$ Department of Biological Sciences, College of Natural Sciences, Redeemer's University; Ede; Osun State; Nigeria

$^{22}$ Center for Food Safety and Applied Nutrition, U.S. Food and Drug Administration; College Park; MD; USA

$^{23}$ Bioinformatics Laboratory National Laboratory of Scientific Computation LNCC/MCTI; Rio de Janeiro; Brazil

@@

The PHA4GE hAMRonization Detection specification was designed to harmonize information prioritised by public health practitioners for genomic surveillance, outbreak investigations and research with direct public health impacts. Fields from various widely used AMR detection tools were compared, and a set of 36 fields were selected to capture information regarding the software and database provenance, information about reference genes and genomes, as well as mutations and resistance genes detected in sequences of interest. The field labels are standardised using terms sourced or contributed to open-source ontologies. Ontologies are sets of controlled vocabulary arranged in a hierarchy, in which terms are linked using logical relationships and the meanings of the terms are disambiguated by the assignment of unique and persistent identifiers. The positions of detected mutations with respect to reference genes/genomes/proteins are specified using the machine-readable Human Genome Variation Society (HGVS) notation system, a sequence variant nomenclature that has been used previously for encoding tuberculosis drug resistance mutations (den Dunnen et al., 2016). To translate HGVS notations into clinician-friendly, human-readable language, a simplified interpretation of mutation coordinates is also provided within the specification. A list of the fields in the specification, along with their ontology identifiers, definitions, expected value types, examples of use, and additional guidance for implementation, are presented in Table \ref{tab:ch6_table1}. It should be noted that the guidance is meant to provide recommendations for developers and may contain suggestions that go beyond the immediate aims of the specification.

\input{tables/chapter_6/table1.tex}

@@

To enable automated data transformations from variable tool outputs to the PHA4GE hAMRonization specification, fields from 18 gene and mutation detection tools (i.e. ABRicate, AMRFinderPlus, AMRPlusPlus, ARIBA, C-SSTAR, DeepARG, GROOT, KmerResistance, ResFams, ResFinder (includes PointFinder), SraX, SRST2, and StarAMR) were mapped to the specification. It’s worth noting that not all tools provide values for all specification fields. Also, many tools provide outputs beyond what is captured by the specification.

Parsers were then constructed to automate the transformation and movement of outputs into the appropriate fields in the hAMRonization reports (Figure \ref{fig:chap6_figure1}). Parser code is open source, and all code and installation instructions are available at \url{https://github.com/pha4ge/hAMRonization}. hAMRonization can be executed from the command-line, and alternatively, can be installed and used in Galaxy via the Galaxy tool shed.
hAMRonization provides both tabular and interactive reporting formats. The interactive report provides the user with an HTML file that can be opened within the browser for navigable data exploration.

\begin{figure*}[]
\centering
\includegraphics[width=\textwidth]{figures/chapter 6/Fig 1.png}
\caption{ Overview of hAMRonization within standard AMR genomics/metagenomics workflows. The hAMRonization module and command-line utility tools allow the creation of a single unified output report from any of 18 established disparate AMR gene and/or variant prediction tools. This facilitates users such as national public health reference laboratories and academic institutions to compare results across tools and change their workflow to different tools without having to develop custom code.  It also means that the communication of results to downstream knowledge users, such as infection control prevention clinicians, can be done in a consistent standardised manner regardless of which AMR prediction tool was used in the genomic/metagenomic analysis.}
\label{fig:chap6_figure1}
\end{figure*}

@@

The developed  hAMRonization snakemake pipeline (see \ref{ch6_materials_workflow}) was used to run two sets of data through a selection of AMR detection tools, using various AMR gene databases.

@@

The ReLAVRA pilot included National Reference Laboratories from Argentina, Brasil, Chile, Colombia, Costa Rica, México and Paraguay. Laboratories and were provided with a dataset of 20 \textit{Shigella sonnei} genomes from a previous Latin America regional project analysed with 5 different AMR detection tools: ABRIcate, AMRFinderPlus, ARIBA, SRST2, StarAMR. Participants received the outputs of the AMR detection tools together with the instructions for installing and running hAMRonization. When participants completed the exercise, they were asked to complete the structured poll. Twelve responses from the seven participating nations were received, and a summary report was prepared by ReLAVRA and provided to PHA4GE. Most of the participants were bioinformaticians and laboratory technicians. For bioinformaticians, the installation instructions were easy to follow, however, laboratory technicians without a bioinformatics background expressed difficulties largely with the dependencies installation. Participants agreed that all harmonised output fields provided relevant information for AMR detection but the most useful were the fields that cover the AMR gene detected and the confidence of that result. Respondents were concerned about the importance of having knowledge of AMR and genomic data analysis for the correct interpretation of the results, especially with regard to decision-makers in clinical settings. An example related to this concern was that when detecting \textit{blaTEM-1} gene by AMRFinderPlus, the field “antimicrobial agent” indicates "Beta-lactam", but this type of drug includes a wide variety of antibiotics, from amoxicillin to extended-spectrum cephalosporins (ESC) and carbapenems. However, \textit{TEM-1} is an extended-spectrum beta-lactamase that does not confer resistance to ESC or carbapenems. While this concern arises from the output of AMRFinderPlus rather than hAMRonization, this challenge is a barrier to the implementation of genomics for AMR surveillance in clinical settings and should be addressed. PHA4GE is cataloguing such critical challenges and will work with the community to solve them. In terms of output summary format, most of the participants preferred the tabular report, citing that a single file combining all the relevant information was more convenient and amenable to processing with different downstream software. Some participants preferred citing its user-friendly visual interface was helpful for comparing which tools identified the same gene.

@@

A dataset of 89 Klebsiella pneumoniae were analysed using RGI, relying on BLASTp and the CARD database, RGI btw, relying on Bowtie2 and the CARD database, AMRFinderPlus which uses BLASTx and the BARRG database, and ResFinder with BLASTn and the ResFinder database. The aim was to determine the number of unique AMR genes identified, with a large variation having been identified across the four tools (Figure \ref{fig:chap6_figure2}). RGI bwt and RGI identified the highest number of hits (n=223 and n=140 respectively), albeit using different identification methods both use the same database. AMRFinderPlus and Resfinder identified an order of magnitude less AMR genes (n=70 and n=46 respectively), even though both rely on BLAST, as does RGI. This highlights the impact of the database chosen when assessing AMR genes.

\begin{figure*}[]
\centering
\includegraphics[width=\textwidth]{figures/chapter 6/Fig 2.png}
\caption{Unique AMR genes idenfiied by RGI, RGI bwt, AMRFInderPlus and ResFinder. The heatmap represents the percenetage of AMR genes found by the combination of the four tools. Although a very large percentage of the ResFinder and AMRFinderPlus detected genes are identified by all tools, RGI and in particular RGI bwt identify a very large subset of AMR genes not present in the othe datasets.}
\label{fig:chap6_figure2}
\end{figure*}

@@

A set of 87 \textit{Pseudomonas aeruginosa} sequences (ST 111) associated with a Metallo-Carbapenemases resistance outbreak in the United Kingdom \cite{witney_genome_2014}, were analysed using ABRIcate, AMRFinderplus, C-SSTAR, Resfinder, StarAMR with default parameters and databases. The resulting outputs were combined using hAMRonization (Figure \ref{fig:chap6_figure3}). Most isolates (n=73) carried the \textit{VIM-2} gene, conferring the observed resistance phenotype. Some genes, such as \textit{crpP} and \textit{sul1},  are repeatedly identified by most of the tools. Differences in gene nomenclature in different databases complicate comparisons e.g. the same gene was also seen with different spelling, such as \textit{blaVIM-2} being called a \textit{blaVIM}. Nomenclature issues aside, harmonised outputs from different tools served to increase the body of evidence (and confidence in results) indicating the role of different AMR determinants in the outbreak.

\begin{figure*}[]
\centering
\includegraphics[width=\textwidth]{figures/chapter 6/Fig 3.png}
\caption{Heatmap of the number of predictions made by the AMR gene detection tools (abricate, amrfinderplus, csstar, resfinder, staramr) that find each resistance gene, overlaid on a tree from those isolates.}
\label{fig:chap6_figure3}
\end{figure*}

@@

\subsection{hAMRonization Proof of Concept Workflow} \label{ch6_materials_workflow}

@@

\section{Supplemental Materials}

\input{tables/chapter_6/suptable1.tex}

\input{tables/chapter_6/suptable2.tex}

### 2022-11-04 chapters/chapter_7.tex
finish hAMRonization paper, fix names formatting on papers

Emma J. Griffiths$^1$, 
Ruth E. Timme$^2$,
Catarina I. Mende$^3$,
Andrew J Page$^4$,
Nabil-Fareed Alikha$^4$,
Dan Fornika$^5$,
Finlay Maguire$^6$,
Josefina Campos$^7$,
Daniel Park$^8$,
Idowu B. Olawoy$^{9,10}$,
Paul E. Oluniy$^{9,10}$,
Dominique Anderson$^{11}$,
Alan Christoffel$^{11}$,
Anders Gonçalves da Silva$^{12}$,
Rhiannon Cameron$^1$,
Damion Dooley$^1$,
Lee S. Katz$^{13}$,
Allison Black$^{14}$,
Ilene Karsch-Mizrach$^{15}$,
Tanya Barret$^{15}$,
Anjanette Johnston$^{15}$,
Thomas R. Connor$^{16,17}$,
Samuel M. Nicholls$^{18}$,
Adam A. Witney$^{19}$,
Gregory H. Tyson$^{20}$,
Simon H. Tausch$^{21}$,
Amogelang R. Raphenya$^{22}$,
Brian Alcock$^{22}$,
David M. Aanensen$^{23,24}$,
Emma Hodcrof$^{25}$,
William W. L. Hsiao$^{1,5,26}$,
Ana Tereza R Vasconcelos$^{27}$,
Duncan R MacCannel$^{28}$,
on behalf of the Public Health Alliance for Genomic Epidemiology (PHA4GE) consortium

>>>>

Emma J. Griffiths$^1$, 
Ruth E. Timme$^2$,
Catarina I. Mende$^3$,
Andrew J. Page$^4$,
Nabil-Fareed Alikha$^4$,
Dan Fornika$^5$,
Finlay Maguire$^6$,
Josefina Campos$^7$,
Daniel Park$^8$,
Idowu B. Olawoy$^{9,10}$,
Paul E. Oluniy$^{9,10}$,
Dominique Anderson$^{11}$,
Alan Christoffel$^{11}$,
Anders Gonçalves da Silva$^{12}$,
Rhiannon Cameron$^1$,
Damion Dooley$^1$,
Lee S. Katz$^{13}$,
Allison Black$^{14}$,
Ilene Karsch-Mizrach$^{15}$,
Tanya Barret$^{15}$,
Anjanette Johnston$^{15}$,
Thomas R. Connor$^{16,17}$,
Samuel M. Nicholls$^{18}$,
Adam A. Witney$^{19}$,
Gregory H. Tyson$^{20}$,
Simon H. Tausch$^{21}$,
Amogelang R. Raphenya$^{22}$,
Brian Alcock$^{22}$,
David M. Aanensen$^{23,24}$,
Emma Hodcrof$^{25}$,
William W. L. Hsiao$^{1,5,26}$,
Ana Tereza R. Vasconcelos$^{27}$,
Duncan R. MacCannel$^{28}$,
on behalf of the Public Health Alliance for Genomic Epidemiology (PHA4GE) consortium
